Model Checking Differentially Private Properties

### 2018-01-30 mdp.tex
proof completed

\begin{theorem}
The $\dpCTLstar$ model checking problem for MDPs  is undecidable.
\end{theorem}
\begin{proof}
The proof follows by a reduction from the \emph{threshold} problem for probabilistic automata.
A \emph{probabilistic automaton}~\cite{Rabin63} is a tuple $\mathcal{A}= (S,\Sigma,M,s_0,F)$ where
\begin{itemize}
  \item $S$ is a finite set of states,
  \item $\Sigma$ is the finite set of input alphabet,
  \item $M: S\times\Sigma\times S \to [0,1]$ such that $\sum_{t\in S} M(s,\alpha,t)=1$ for all $s\in S$ and $\alpha\in \Sigma$,
  \item $s_0\in S$ is the initial state,
  \item $F\subseteq S$ is a set of accepting states.
\end{itemize}

Each input alphabet $\alpha$ induces a stochastic matrix $M(\alpha)$ in the obvious way. Let $\lambda$ denote the empty string. For $\eta\in\Sigma^*$ we define $M(\eta)$ inductively by: $M(\lambda)$ is the identity matrix, $M(x\eta')=M(x)M(\eta')$. Thus, $M(\eta)(s,s')$ denotes the probability of going from $s$ to $s'$ after reading $\eta$. Let $v_F$ denote the characteristic row vector for the set $F$, and $v_{s_0}$ denote the characteristic row vector for the set $\{s_0\}$. Then, the accepting probably of $\eta$ by $\mathcal{A}$ is defined as $v_{s_0}\cdot M(\eta)\cdot (v_F)^c$ where $(v_F)^c$ denotes the transpose of $v_F$. The following \emph{emptyness problem} is know to be undecidable~\cite{Paz71}
%~\cite{Worrell08}:

@@

 For the other direction, 
\end{proof}

For instance, we consider the formula $\phi:=F B$ with $B\subseteq S$ and assume that states in $B$ are absorbing (with only self-loops). For the case $\epsilon=0$, the condition reduces to
$\myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{F B} -
\myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{F B}\le \delta$. If $\delta=0$ it is the classical language equivalence problem for probabilistic automata~\cite{Rabin63}, which can be solved in polynomial time \cite{Tzeng92}. However, if $\delta>0$, the problem
becomes an approximate version of the language equivalence problem~\cite{Tzeng92}. To the best of our knowledge, the decision problem is still open, and has only be investigated for the special case when all states are connected~\cite{Tzeng92}. If in addition $\epsilon>0$, it seems to be a problem that has not been considered in the MDP model checking community before. Thus we leave the general model checking algorithm as future works. \lz{see this}

\hide{
Since the scheduler attaining the maximally probable
behavior from a state may be different from the scheduler attaining
the maximally probable behavior from its neighbors, two
neighbors may still be distinguished by a scheduler. The
weaker definition does not preserve differential privacy. We hence
prefer the original definition.
}

>>>>

\begin{theorem}
The $\dpCTLstar$ model checking problem for MDPs  is undecidable.
\end{theorem}
\begin{proof}
The proof follows by a reduction from the \emph{emptiness} problem for probabilistic automata.
A \emph{probabilistic automaton}~\cite{Rabin63} is a tuple $\mathcal{A}= (S,\Sigma,M,s_0,F)$ where
\begin{itemize}
  \item $S$ is a finite set of states,
  \item $\Sigma$ is the finite set of input alphabet,
  \item $M: S\times\Sigma\times S \to [0,1]$ such that $\sum_{t\in S} M(s,\alpha,t)=1$ for all $s\in S$ and $\alpha\in \Sigma$,
  \item $s_0\in S$ is the initial state,
  \item $F\subseteq S$ is a set of accepting states.
\end{itemize}

Each input alphabet $\alpha$ induces a stochastic matrix $M(\alpha)$ in the obvious way. Let $\lambda$ denote the empty string. For $\eta\in\Sigma^*$ we define $M(\eta)$ inductively by: $M(\lambda)$ is the identity matrix, $M(x\eta')=M(x)M(\eta')$. Thus, $M(\eta)(s,s')$ denotes the probability of going from $s$ to $s'$ after reading $\eta$. Let $v_F$ denote the characteristic row vector for the set $F$, and $v_{s_0}$ denote the characteristic row vector for the set $\{s_0\}$. Then, the accepting probably of $\eta$ by $\mathcal{A}$ is defined as $v_{s_0}\cdot M(\eta)\cdot (v_F)^c$ where $(v_F)^c$ denotes the transpose of $v_F$. The following \emph{emptiness problem} is know to be undecidable~\cite{Paz71}
%~\cite{Worrell08}:

@@

 For the other direction, assume that all $\eta\in\Sigma^*$ it holds $v_{s_0}\cdot M(\eta)\cdot (v_F)^c \leq 0$. For the sake of contradiction, assume that $s_0\not\models \dpriv{1}{0} (\diamond at_F)$. Since the relation $N_S$ contains only one pair, it implies that there exists a scheduler $\scheduler{S}\in\Sigma^\omega$ such that $\myPr{s_0}{M_{\scheduler{S}}}{\neighbor{S}}{\diamond at_F} >0$. It is then easy to construct a finite sequence $\eta\in\Sigma^*$ with  $v_{s_0}\cdot M(\eta)\cdot (v_F)^c > 0$, a contradiction.
\end{proof}

We consider some decidable special cases. Consider the formula $\phi:=F B$ with $B\subseteq S$ and assume that states in $B$ are absorbing (with only self-loops). For the case $\epsilon=0$, the condition reduces to
$\myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{F B} -
\myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{F B}\le \delta$. If $\delta=0$ it is the classical language equivalence problem for probabilistic automata~\cite{Rabin63}, which can be solved in polynomial time \cite{Tzeng92}. However, if $\delta>0$, the problem
becomes an approximate version of the language equivalence problem~\cite{Tzeng92}. To the best of our knowledge, the decision problem is still open, and has only be investigated for the special case when all states are connected~\cite{Tzeng92}. 
p
\hide{
Since the scheduler attaining the maximally probable
behavior from a state may be different from the scheduler attaining
the maximally probable behavior from its neighbors, two
neighbors may still be distinguished by a scheduler. The
weaker definition does not preserve differential privacy. We hence
prefer the original definition.
}

### 2018-01-30 specifications.tex
proof completed

\paragraph{Reachability Privacy.}
For our threshold \lz{to do} model, we have the differential privacy property $\wedge_{i=1}^{c} \dpriv{\epsilon}{0}(\X^k \top) \wedge \dpriv{\epsilon}{0}(F \top)$. Here $\dpriv{\epsilon}{0}(F \top)$ is a reachability privacy: it states that related states should have similar probabilities of eventually reaching $\top$.

>>>>

\paragraph{Reachability Privacy.}
The formula $\dpriv{\epsilon}{\delta}(F \top)$ is a reachability privacy: it states that related states should have similar probabilities of eventually reaching $\top$. Note neighbor states may reach the goal state $\top$ with

For our threshold \lz{to do} model, we have the differential privacy property $\wedge_{i=1}^{c} \dpriv{\epsilon}{0}(\X^k \top) \wedge \dpriv{\epsilon}{0}(F \top)$.
